Cassini State Capture

### 2019-10-15 writeups/Su_disk_disp.tex
new adiabaticity criterion?

Finally, the Hamiltonian corresponding to the equation of motion
\autoref{eq:dsdt_base} is
\begin{equation}
    H\p{\phi, \theta} = -\frac{1}{2}\p{\hat{s} \cdot \hat{l}}^2
        + \eta \p{\hat{s} \cdot \hat{l}_d}.\label{eq:H}
\end{equation}
Here, we adopt a spherical coordinate system where $\hat{l} = \hat{z}$ and
$\theta, \phi$ are the polar and azimuthal angle of $\hat{s}$. By convention, we
choose $\hat{l}_z$ at coordinates $\theta = I, \phi = \pi$.

@@

Spin states satisfying $\rd{\hat{s}}{t} = 0$ are referred to as \emph{Cassini
States} (CS). When $\eta < \eta_c$, there are four CSs, and when $\eta > \eta_c$
there are only two; $\eta_c$ is \citep{henrard1987,ward2004I}
\begin{equation}
    \eta_c \equiv \p{\sin^{2/3}I + \cos^{2/3}I}^{3/2}.
\end{equation}
CSs 1, 2, 3 are stable while CS4 is unstable.

The locations of the $\cos\theta$ of the two/four CSs as a function of $\eta$ is
provided in \autoref{fig:cs_locs}.
\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{../initial/99_misc/2_cs_locs.png}
    \caption{$\theta$ values of the Cassini states over time, following the same
    coloring scheme as \autoref{fig:eq_1contours}. Note that $\theta \in [-\pi,
    \pi]$ is the traditional definition of the polar angle
    \citep[see e.g.][]{colombo1966,peale1969,henrard1987}.}\label{fig:cs_locs}
\end{figure}

@@

In the four-CS regime, one of the CSs is a saddle point, conventionally denoted
Cassini State 4 (CS4). All trajectories are periodic with finite period except
two critical trajectories asymptotic in the past and future to CS4. Together,
these two critical trajectories are referred to as the \emph{separatrix} and
divide phase space into three zones. The separatrix and three zones are most
clearly shown via a contour plot of $H\p{\phi, \cos \theta}$ in
\autoref{fig:eq_1contours}.
\begin{figure*}
    \centering
    \includegraphics[width=0.8\textwidth]{../initial/0_eta/1contours_flip.png}
    \caption{Contour plot of $H\p{\phi, \cos \theta}$, where warmer colors
    denote more positive values. The black solid line is the separatrix, which
    only exists for $\eta < \eta_c$. The three zones, divided by the separatrix,
    are labeled. The interior of the separatrix, shaded in grey, is formally
    only defined for $\eta < \eta_c$, but we may identify the phase space
    trajectories that flow into zone II when evolved forward in time; this is
    the shaded region in the top left plot, bounded by the dotted black
    line.}\label{fig:eq_1contours}
\end{figure*}

As trajectories of constant $\eta$ evolve along
contours of constant $H$, it can be observed that trajectories in zone II
librate about CS2 while those in zones I/III circulate.

The unsigned areas of the three zones are known exactly in literature
\citep{henrard1987,ward2004I}. Defining
\begin{align*}
    z_0 &= \eta\cos I, &
    \chi &= \sqrt{-\frac{\tan^3\theta_4}{\tan I} - 1},\\
    \rho &= \chi \frac{\sin^2 \theta_4\cos \theta_4}{
        \chi^2 \cos^2\theta_4 + 1},&
    T &= 2\chi \frac{\cos \theta_4}{
        \chi^2 \cos^2\theta_4 - 1},\\
\end{align*}
the areas for $\eta < \eta_c$ are given by
\begin{subequations}\label{se:area_ward}
    \begin{align}
        A_{II} &= 8\rho + 4\arctan T - 8z_0 \arctan \frac{1}{\chi},\\
        A_I &= 2\pi\p{1 - z_0} - \frac{A_2}{2},\\
        A_{III} &= 2\pi\p{1 + z_0} - \frac{A_2}{2}.
    \end{align}
\end{subequations}
These are plotted as a function of $\eta$ in \autoref{fig:eq_areas}. Note that
the zones are not formally defined for $\eta > \eta_c$, but a natural extension
exists: evolve an initial point $p$ under adiabatic decrease of $\eta > \eta_c$
until the separatrix appears at $\eta = \eta_c$, then identify $p$ with the zone
it is in at $\eta_c$. Since phase space area is conserved under adiabatic
evolution, this implies $A_i\p{\eta > \eta_c} = A_i(\eta_c)$.
\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{../initial/99_misc/1_areas.png}
    \caption{Plot of fractional $A_{i}(\eta)$ as given by
    \autoref{se:area_ward}. Dotted lines correspond to small $\eta$
    approximations discussed in \autoref{ss:app_transition}. Dashed lines denote
    effective values of $A_{II}, A_{III}$ for $\eta > \eta_c$, denoting the
    area of points that would flow into either area under adiabatic decrease of
    $\eta$ to below $\eta_c$.}\label{fig:eq_areas}
\end{figure}

@@

One criterion for adiabaticity is that the separatrix crossing timescale be
longer than the the resonant libration period \citep{ward2004I}. This is
evaluated in~\cite{millholland_disk}, and in our problem reduces to
\begin{equation}
    \at{\rd{\eta}{t}}_{cross} = -\epsilon\eta_\star \gtrsim
        -\sin\theta \sin I \cos I.\label{eq:ad_constr}
\end{equation}
Note that this is a necessary but not sufficient criterion: the libration period
is longer for trajectories nearer the separatrix and diverges at the separatrix
for constant $\eta$.

@@

We first seek a simple parameterization for the separatrix. We first solve for
equilibria of the equation of motion \autoref{eq:dsdt_base} to compute the
coordinates for Cassini State 4:
\begin{equation}
    \cos \theta_4 \approx \frac{\mu \cos I}{1 - \eta \sin I}.
\end{equation}
Note that $\phi_4 = 0$. Then, the separatrix is the level curve of the
Hamiltonian intersecting CS4, so it satisfies $H\p{\theta_{sep}(\phi), \phi} =
H\p{\theta_4, \phi_4}$. This solves to be approximately
\begin{equation}
    \cos \theta_{sep}(\phi) \approx \cos \theta_4 \pm
        \sqrt{2\eta \sin I\p{1 - \cos \phi}}.
\end{equation}
Integration of the phase area enclosed by the two legs of the separatrix then
yields
\begin{equation}
    A_{II}(\eta) \approx 16\sqrt{\eta \sin I}.\label{eq:a_approx}
\end{equation}
This, in conjunction with \autoref{eq:henrard_hop}, is sufficient to compute
$\theta_f\p{\theta_{sd, i}}$ for zone II initial conditions.

>>>>

Finally, the Hamiltonian corresponding to the equation of motion
\autoref{eq:dsdt_base} is
\begin{equation}
    \mathcal{H}\p{\theta, \phi} = -\frac{1}{2}\p{\hat{s} \cdot \hat{l}}^2
        + \eta \p{\hat{s} \cdot \hat{l}_d}.\label{eq:H}
\end{equation}
Here, we adopt a spherical coordinate system where $\hat{l} = \hat{z}$ and
$\theta, \phi$ are the polar and azimuthal angle of $\hat{s}$. By convention, we
choose $\hat{l}_z$ at coordinates $\theta = I, \phi = \pi$, in which case the
Hamiltonian can be written
\begin{equation}
    \mathcal{H} = -\frac{1}{2}\cos^2\theta
        + \eta \p*{\cos \theta \cos I - \sin I \sin \theta \cos \phi}.
\end{equation}

Note that $\p{\cos \theta, \phi}$ form a canonically conjugate pair of variables
describing $\hat{s}$. The equations of motion for these variables can be derived
by taking derivatives of $\mathcal{H}$ given by \autoref{eq:H} and agree with
\autoref{eq:dsdt_base}:
\begin{subequations}\label{se:H_eom}
    \begin{align}
        \pd{\phi}{t} = \pd{\mathcal{H}}{(\cos\theta)}
            &= -\cos\theta + \eta\p{\cos I + \sin I \cot \theta \cos \phi},\\
        \pd{(\cos \theta)}{t} = -\pd{\mathcal{H}}{\phi}
            &= -\eta \sin I \sin \theta \sin \phi.
    \end{align}
\end{subequations}
While these equations are numerically stiff owing to the $\cot\theta$ term, they
are the most intuitive description for analytical work.

@@

Spin states satisfying $\rd{\hat{s}}{t} = 0$ are referred to as \emph{Cassini
States} (CSs). There are either two or four solutions, depending on the values
of $\eta, I$. Following standard convention and nomenclature (see
\autoref{fig:cs_locs}), CSs 1, 3, 4 have $\phi = 0, \theta < 0$ and CS2 has
$\phi = \pi, \theta > 0$. Of these, CSs 1, 2, 3 are stable while CS4 is
unstable.

When $\eta < \eta_c$, where
\begin{equation}
    \eta_c \equiv \p{\sin^{2/3}I + \cos^{2/3}I}^{-3/2},
\end{equation}
all four CSs exist, and when $\eta > \eta_c$, only CSs 2, 3 exist
\citep{henrard1987,ward2004I}. The CS obliquities as a function of $\eta$ are
provided in \autoref{fig:cs_locs}.
\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{../initial/99_misc/2_cs_locs.png}
    \caption{$\theta$ values of the Cassini states over time, following the same
    coloring scheme as \autoref{fig:eq_1contours}. Note that $\theta \in [-\pi,
    \pi]$ is the traditional definition of the polar angle
    \citep[see e.g.][]{colombo1966,peale1969,henrard1987}.}\label{fig:cs_locs}
\end{figure}

@@

When $\eta < \eta_c$, CS4 exists and is a saddle point. All trajectories are
periodic with finite period except two critical trajectories asymptotic in the
past and future to CS4. Together, these two critical trajectories are referred
to as the \emph{separatrix} and divide phase space into three zones. The
separatrix, three zones, and their relations to the CSs are most clearly shown
via a contour plot of $\mathcal{H}\p{\phi, \cos \theta}$ in
\autoref{fig:eq_1contours}.
\begin{figure*}
    \centering
    \includegraphics[width=0.8\textwidth]{../initial/0_eta/1contours_flip.png}
    \caption{Contour plot of $\mathcal{H}\p{\phi, \cos \theta}$, where warmer
    colors denote more positive values. The black solid line is the separatrix,
    which only exists for $\eta < \eta_c$. The three zones, divided by the
    separatrix, are labeled. The interior of the separatrix, shaded in grey, is
    formally only defined for $\eta < \eta_c$, but we may identify the phase
    space trajectories that flow into zone II when evolved forward in time; this
    is the shaded region in the top left plot, bounded by the dotted black
    line.}\label{fig:eq_1contours}
\end{figure*}

As trajectories of constant $\eta$ evolve along contours of constant
$\mathcal{H}$, it can be observed that trajectories in zone II librate about CS2
while those in zones I/III circulate.

The unsigned areas of the three zones are known exactly
\citep{henrard1987,ward2004I}. Defining
\begin{align*}
    z_0 &= \eta\cos I, &
    \chi &= \sqrt{-\frac{\tan^3\theta_4}{\tan I} - 1},\\
    \rho &= \chi \frac{\sin^2 \theta_4\cos \theta_4}{
        \chi^2 \cos^2\theta_4 + 1},&
    T &= 2\chi \frac{\cos \theta_4}{
        \chi^2 \cos^2\theta_4 - 1},\\
\end{align*}
the areas for $\eta < \eta_c$ are given by
\begin{subequations}\label{se:area_ward}
    \begin{align}
        A_{II} &= 8\rho + 4\arctan T - 8z_0 \arctan \frac{1}{\chi},\\
        A_I &= 2\pi\p{1 - z_0} - \frac{A_2}{2},\\
        A_{III} &= 2\pi\p{1 + z_0} - \frac{A_2}{2}.
    \end{align}
\end{subequations}
These are plotted as a function of $\eta$ in \autoref{fig:eq_areas}. Note that
the zones are not formally defined for $\eta > \eta_c$, but a natural extension
exists: evolve an initial point $p$ under adiabatic decrease of $\eta > \eta_c$
until the separatrix appears at $\eta = \eta_c$, then identify $p$ with the zone
it is in at $\eta_c$. Since phase space area is conserved under adiabatic
evolution, this implies $A_i\p{\eta > \eta_c} = A_i(\eta_c)$.
\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{../initial/99_misc/1_areas.png}
    \caption{Plot of fractional $A_{i}(\eta)$ as given by
    \autoref{se:area_ward}. Dotted lines correspond to small $\eta$
    approximations discussed in \autoref{ss:app_transition}. Dashed lines denote
    effective values of $A_{II}, A_{III}$ for $\eta > \eta_c$, denoting the
    area of points that would flow into either area under adiabatic decrease of
    $\eta$ to below $\eta_c$.}\label{fig:eq_areas}
\end{figure}

@@

One necessary criterion for adiabaticity is that the separatrix crossing
timescale be longer than the the resonant libration period \citep{ward2004I}.
The resonant libration period is just the libration period about CS2, which can
be found by linearizing \autoref{se:H_eom} about $\phi \approx \pi, \theta
\approx \theta_2$. The resulting criterion is
\begin{align}
    -\at{\frac{1}{\eta}\rd{\eta}{\tau}}_{cross} = \epsilon &\lesssim
            \frac{2\pi}{T_{lib}},\\
        &\lesssim \sqrt{\eta_\star\sin\theta \sin I
            \p{1 + \eta_\star \frac{\sin I}{\sin^3\theta}}}.\label{eq:ad_constr}
\end{align}
Note that this is a necessary but not sufficient criterion: the libration period
is longer for trajectories nearer the separatrix and diverges at the separatrix
for constant $\eta$. This formula also is in slight disagreement with that
presented in \citealt{millholland_disk}

@@

We first seek a simple parameterization for the separatrix. We first solve for
equilibria of the equation of motion \autoref{eq:dsdt_base} to compute the
coordinates for Cassini State 4:
\begin{equation}
    \cos \theta_4 \approx \frac{\mu \cos I}{1 - \eta \sin I}.
\end{equation}
Note that $\phi_4 = 0$. Then, the separatrix is the level curve of the
Hamiltonian intersecting CS4, so it satisfies $\mathcal{H}\p{\theta_{sep}(\phi),
\phi} = \mathcal{H}\p{\theta_4, \phi_4}$. This solves to be approximately
\begin{equation}
    \cos \theta_{sep}(\phi) \approx \cos \theta_4 \pm
        \sqrt{2\eta \sin I\p{1 - \cos \phi}}.
\end{equation}
Integration of the phase area enclosed by the two legs of the separatrix then
yields
\begin{equation}
    A_{II}(\eta) \approx 16\sqrt{\eta \sin I}.\label{eq:a_approx}
\end{equation}
This, in conjunction with \autoref{eq:henrard_hop}, is sufficient to compute
$\theta_f\p{\theta_{sd, i}}$ for zone II initial conditions.
